### 2012-07-05 paper/chemokine.tex
Lots of minor edits to figures and text

The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.  Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes \cite{Maines2008, Saenz2010, Hatta2010}, temporal and spatial details of precursor T and B cell activation within the lymph node architecture [Legge and Braciale 2006], and functional capabilities of resulting effector lymphocytes [La Gruta and Doherty 2006].  In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora \cite{Marshall2001}.

@@

Here we focus on the time and physical constraints of the lymphocyte diaspora.  How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control \cite{Allan1990} [La Gruta and Doherty 2006], after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?  A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Kawai1999, Sallusto2000, Zhao2000, Dufour2002, Thelen2008, Sigmundsdottir2008, Sallusto2008, Bromley2008, Kohlmeier2008}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine local concentration, function and efficiency are difficult to examine in a virus-infected whole animal model.

@@

@@

% FIXME: The main goal of this paper is to compare chemokine signaling effectiveness between infections from different strains of influenza. Our study looks at separate infections from three different strains: seasonal, avian, and swine. Futhermore, we compare the roles of two established signaling molecules: CXCL10 (IP-10) and CCL5 (RANTES).

FIXME: Need to say somewhere that we simulate a generic chemokinetic signal that abstracts X, Y, and Z.

Computational modeling was conducted using Cycells [REF], a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response [REF]. A simplified model of T-Cell activation and recirculation was implemented in Cycells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions. To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T-cells descend through the vascular tree until chemotactic signal is detected, at which point they exit the vasculature and follow the [FIXME chemotactic/inflammatory] gradient to the site of infection. T-cells that do not encounter chemokine recirculate to the lymph node. Once at the site of infection, a T-Cell may encounter an infected epithelial cell and induce apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. The chemokine diffuses from secreting cells, thus increasing the size of the signaling area. After a delay, to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue. Because T-Cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations. Figure \ref{fig:modelchart} illustrates the model components 
(epithelial cells, T-cells, and particles) and how they transition between different states.

@@

\begin{table}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion & .0318 & Stokes-Einstein / Beau \\
  Viral Decay &  1/day & Soumya \\
  Chemokine Diffusion & .318 & Stokes-Einstein \\
  Chemokine Decay &  3.8508e-4/s & 30m half-life \\
  Infection Sensitivity &  2h/virion & [1] \\
  Incubation Time &  10h & Mitchell \\
  Expression Time &  16.7h & Mitchell \\
  Apoptosis Time & 1h & Fred \\
  T-Cell Production Rate & 777/h & Soumya, Miao fits \\ 
  Circulation Time & 6m & Fred, mouse model \\
  Search Time & 10m & Fred, mouse model \\
  T-Cell Speed (Search) & $30 \mu m/s$ & Fred ? \\
  T-Cell Speed (Activated) & $3 \mu m/s$ & Fred ?\\
  T-Cell Sensitivity & $100 ng/ml$ & See Results \\
  T-Cell Expected Kill Rate & 10m & Fred/Drew \\
  Cell Radius & $25 \mu m$ & Fred \\
  T-Cell Age (Active) & 2h & Fred \\
  T-Cell Age (Other) & 3d & Fred \\
  T-Cell Ramp Up Time & 5d & Icaris \\
  IgM Strength & Viral decay of 3/day after day 3 & Soumya [3] \\
  \hline  
\end{tabular}
\caption{Generic model parameters}
\label{table:parameters}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ | r | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  IP-10 Production Rate &  & [2]\\
  Avian H5N1 & 7.2e-5 ng/s$\cdot$cell & \\
  Seasonal H1N1 & 9.3e-5 ng/s$\cdot$cell& \\
  Pandemic H1N1 & 6.3e-5 ng/s$\cdot$cell& \\
  \hline
  RANTES Production Rate & & [2] \\
  Avian H5N1 & 3.9e-5 ng/s$\cdot$cell& \\
  Seasonal H1N1 & 1.3e-6 ng/s$\cdot$cell& \\
  Pandemic H1N1 & 1.4e-6 ng/s$\cdot$cell& \\
  \hline
  Virus Production Rate &  & Mitchell \\
  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell& \\
  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell& \\
  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell& \\  
  \hline  
\end{tabular}
\caption{Strain specific model parameters}
\label{table:strains}
\end{center}
\end{table}

\begin{itemize}
\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/ml.  Thus, if you multiply the local viral concentration by 9.05e12 you get an infection event after two hours given that concentration.  This probability scales linearly with the viral concentration.
\item[2] These are calculated off of Fred's new data sets using the Mitchell ODE model and parameters as a basis.
\item[3] IgM is represented by increasing the viral decay rate to 3/day after the third day of the simulation.
\end{itemize}

@@

Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells.  Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm [REF: GA-ref].  The model estimates are shown in Table~\ref{table:strains}.  These chemokine expression rates are consistent for each chemokine across the different strains except that the RANTES rate is higher in avian H5N1.  There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{data}
 \end{center}
\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).  For each strain, the viral load (blue circle) is shown in PFU/ml and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/ml.} 
 \label{fig:data}
\end{figure}

@@

% [FIXME-DREW: Shouldn't we do 30 runs instead of 10 to get statistical significance?   Drew: Sure, I can always do more runs, I did 10 just to have a figure to look at.]

Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).  To test the strength of this effect, we ran each model ten (MORE?) times using the default parameters given in Table~\ref{table:parameters}, finding that the variance of infected cells was X.  Each run took as input the calculated viral production and chemokine production rates for the three different strains of influenza, reporting as output the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.  The figures thus approximate plaque growth over time.

@@

The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).  The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.  By the time T-Cells arrive, the plaque is too large for them  to cover effectively.  Because pH1N1 is produced so quickly, the window of opportunity for T-Cells to gain ground is too small (Discussion).  aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{variance}
 \end{center}
\caption{Model results: Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections. IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.  The ten runs were initialized identically for each strain.  Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/ml).} 
 \label{fig:variance}
\end{figure}

@@

T-Cells exit a capillary when they sense inflammatory cytokine and then follow a second signal, the chemokine gradient.  We use the chemokine gradient as a proxy for the inflammatory signal.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.

To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude. We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e-18 and 1e-19 (1e-19 and 1e-20 for pH1N1), above which there is no discernible effect on infection kinetics.  Because multiple T-cells in the same area do not increase the rate at which  apoptosis is induced, we hypothesize that a critical number of T-cells is required to clear an infection, above which there is no additional benefit from increasing numbers.  For consistency, we standardized all subsequent model runs to  sensitivity of 1e-22.

@@

Because aH5N1 has been shown to surpress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 is ineffectual (FIXME: need to clean this up).  This may explain why aH5N1 has a much higher secretion rate for RANTES than the other two strains (Table~\ref{table:strains}).  Because of this behavior, IP-10 was not included in the aH5N1 models.

Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the efficacy of the simulated immune response (Fig.~\ref{fig:chemokine}).  The lack of both chemokines leads to runaway infections in all three strains.  The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.  IP-10 alone proves to be just as effective and both IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.

@@

Early in the infection (up to day 4) the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare. Over time, cells in the plaque's interior die, so active cells comprise a decreasing proportion of the total. T-cells arrive at day 5 and begin to killing the plentiful virus-expressing cells. By day 6 many of the expressing cells have been eliminated and the plaque is dominated by dead cells.  In the H5N1 example, the plaque is dense, so the secreting cells can be found easily and the infection is eliminated.  In both H1N1 simulations, however, secreting cells are not eliminated.  In these cases, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.  T-cells still accumulate, but they arrive at a slower rate than the plaque's rate of growth. This leads to a lower probability of T-cells finding and killing the remaining expressing cells, and the infection is able to recover and resume growth.   The regions of concentrated chemokine trail the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine and for old pockets of chemokine to decay away.  Thus, T-cells, which rely on the chemokine gradient to find infected cells, often lag behind changes in the plaque.  The delayed response and the low proportion of virus secreting cells prevent T-cells from clearing the infection completely.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{cycells}
 \end{center}
\caption{Example simulated sH1N1 infection. Screenshots taken at Day 4, Day 5.5, and Day 7.  The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).  Free virus and chemokine particles are represented by compartmentalized concentrations of mols/ml and ng/ml (see the legend in the bottom right).} 
 \label{fig:cycells}
\end{figure}

@@

The immune response fails to clear sH1N1 completely and is unable to contain the pH1N1 (Fig. \ref{fig:variance}).  This appears to be an artifact of the spatial nature of the ABM.  Why can the immune response create an immediate decrease in the number of pH1N1 infected cells at day 5, only to fail in the same endeavor two days later?   Figure \ref{fig:variance} reports only the number of infected cells, not counting dead cells in the focus of infection (FOI).  Further, T-Cells induce apoptosis in virus-secreting cells, but not in virus-incubating cells.  At day 5.5, the ratio of secreting cells to the total FOI is high (Fig.~\ref{fig:cycells}). By day 7, the ratio of plaque size to the number of secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.  However the greater replication rate of pH1N1 widens the gap (Fig.~\ref{fig:plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to keep up, even after day seven, leading to control (but not elimination) of the sH1N1 infection.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s [FIXME: Where does this come from? IS this our data, model results, or what? Drew: from Mitchell et al].  That is, a new viral particle is produced approximately every 200 seconds.  Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit a single infected cell to producing 18 new viral particles.  Without additional immune responses, T-Cells alone cannot contain the pH1N1 infection.  In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, so the T-cell can limit a single infected cell to producing 1.3 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=0.7\textwidth]{tcellratio}
 \end{center}
\caption{A logarithmic plot of the ratio of T-Cells at the focus of infection vs the number of cells that make up the focus of infection.  aH5N1 exhibits unstable behavior once the infection is cleared and T-cell counts are reduced to negligible levels.} 
 \label{fig:tcellratio}
\end{figure}

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{plaquesize}
 \end{center}
\caption{Comparison of total plaque size (green), number of virus incubating cells (blue), number of virus secreting cells (red), and T-cells (indigo) for the simulated infections of H5N1, sH1N1, and pH1N1.  NOTE TO SELF: Text came out tiny!  Legend is outdated.} 
 \label{fig:plaquesize}
\end{figure}

@@

Another advantage of our ABM is the ability to render the model using OpenGL (Fig.~\ref{fig:cycells}).  The ability to see the model in action reveals spatial dynamics that are absent in ODE models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems.  For example, observing the simulation reveals challenges to T-cell detection of infected cells that are not apparent using standard population plots.  By watching the model progress, we observed three key phenomena.  First, because T-cells find infected cells by climbing the chemokine gradient, we see T-cell clustering at local maxima of chemokine concentration.  Thus, we have an explanation for why T-cells do not increase in effectiveness as their numbers increase.  Second, we notice that clumps of T-cells remain in place even after all virus-secreting cells have been killed off.  This is because it takes time for the local chemokine maxima to diffuse and decay to a point where the T-cells can once again climb the gradient to a new maxima.  Finall, we can see that infected cells grow more disperse as the size of the infection grows.  Because T-cells cluster together they become incapable of covering the increasing plaque size.  These spatial observations provide explanations for the pH1N1 resurgence that would be hidden without the visualization tools provided by the ABM.

@@

In both H1N1 strains the T-cell resposne fails to clear the infection by day 10.  In the case of pandemic H1N1, the infection continues to accelerate after a short period of clearance.  We suggest that this is due to the absence of several innate immune agents that were not included in the model.  We chose to remove most of the innate response in order to better focus on the relative effects of the cytokine signaling and T-cell response.  Although the infections do not clear by day 10, the models show clear differences in the ability of the immune system to deal with influenza strains of varying replication rates.  We see that the viral production rate is a major factor in the strength of the infection.

>>>>

The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.  Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes \cite{Maines2008, Saenz2010, Hatta2010}, temporal and spatial details of precursor T and B cell activation within the lymph node architecture \cite[Chapter 5]{Wormser2006a}, and functional capabilities of resulting effector lymphocytes \cite[Chapter 6]{Wormser2006a}.  In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora \cite{Marshall2001}.

@@

Here we focus on the time and physical constraints of the lymphocyte diaspora.  How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control \cite[Chapter 6]{Wormser2006a}\cite{Allan1990}, after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?  A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Kawai1999, Sallusto2000, Zhao2000, Dufour2002, Thelen2008, Sigmundsdottir2008, Sallusto2008, Bromley2008, Kohlmeier2008}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine local concentration, function and efficiency are difficult to examine in a virus-infected whole animal model.

@@

We hypothesize that the variation of virulence between the three strains of influenza arise from deviations in viral production and chemokine production from infected epithelial cells as well as differences in the local interactions of T-cells with infected cells and chemokines.  We find that while viral production does vary greatly between strains, chemokine production does not.  Two chemokines show increased production during infections: CXCL10 (IP-10) and CCL5 (RANTES).  Of the two, IP-10 consistently shows higher per cell production across all strains while RANTES is only produced at a comparable rate by aH5N1 influenza, which we hypothesize is in compensation for the inability of IP-10 to function in the absence of interferon (which is suppressed by aH5N1 \cite{Mitchell2011}).  Furthermore, the spatial nature of our model reveals difficulties in T-cell search not exposed in spatially homogeneous ODE models.  We see that an increase in the number of T-cells at the focus of infection past a certain threshold does not improve the rate of clearance of the infection and that T-cells face an increasing challenge as plaque sizes increase in diameter over the course of an infection.

@@

Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response. A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions. To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T-cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T-cells that do not encounter cytokine recirculate to the lymph node. Once at the site of infection, a T-Cell may encounter an infected epithelial cell and induce apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. The chemokine diffuses from secreting cells, thus increasing the size of the signaling area. After a delay to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue. Because T-Cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations. Figure \ref{fig:modelchart} illustrates the model components 
(epithelial cells, T-cells, and particles) and how they transition between different states.

@@

\begin{table}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
  Viral Decay &  $1/day$ & Soumya \\
  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
  Chemokine Decay &  $3.8508e-4/s$ & 30m half-life \\
  Infection Sensitivity &  $2 h/virion$ & [1*] \\
  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
  Apoptosis Time & $1 h$ & Fred \\
  T-Cell Production Rate & $777/h$ & Soumya, Miao fits \\ 
  Circulation Time & $6 m$ & Fred, mouse model \\
  Search Time & $10 m$ & Fred, mouse model \\
  T-Cell Speed (Search) & $30 \mu m/s$ & Fred ? \\
  T-Cell Speed (Activated) & $3 \mu m/s$ & Fred ?\\
  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
  T-Cell Expected Kill Rate & $10 m$ & Fred/Drew \\
  Cell Radius & $25 \mu m$ & Drew? \\
  T-Cell Age (Active) & $2 h$ & Fred \\
  T-Cell Age (Other) & $3 d$ & Fred \\
  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
  IgM Strength & Viral decay of $3/day$ after day 3 & Soumya [3*] \\
  \hline  
\end{tabular}
\caption{Generic model parameters}
\label{table:parameters}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ | r | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  IP-10 Production Rate &  & [2*]\\
  Avian H5N1 & 7.2e-5 ng/s$\cdot$cell & \\
  Seasonal H1N1 & 9.3e-5 ng/s$\cdot$cell& \\
  Pandemic H1N1 & 6.3e-5 ng/s$\cdot$cell& \\
  \hline
  RANTES Production Rate & & [2*] \\
  Avian H5N1 & 3.9e-5 ng/s$\cdot$cell& \\
  Seasonal H1N1 & 1.3e-6 ng/s$\cdot$cell& \\
  Pandemic H1N1 & 1.4e-6 ng/s$\cdot$cell& \\
  \hline
  Virus Production Rate &  & \cite{Mitchell2011} \\
  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell& \\
  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell& \\
  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell& \\  
  \hline  
\end{tabular}
\caption{Strain specific model parameters}
\label{table:strains}
\end{center}
\end{table}

\begin{itemize}
\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/mL.  Thus, if you multiply the local viral concentration by 9.05e12 you get an infection event after two hours given that concentration.  This probability scales linearly with the viral concentration.
\item[2] These are calculated off of Fred's new data sets using the Mitchell ODE model and parameters as a basis.
\item[3] IgM is represented by increasing the viral decay rate to 3/day after the third day of the simulation.
\end{itemize}

@@

Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells.  Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.  The model estimates are shown in Table~\ref{table:strains}.  These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.  There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{data}
 \end{center}
\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).  For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.  sH1N1 IP-10 registered three values above the measurement threshold of 8,500 $ng/mL$ (dotted line).  These values were not included in model fitting.} 
 \label{fig:data}
\end{figure}

@@

Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).  To test the strength of this effect, we ran each model fifty times using the default parameters given in Table~\ref{table:parameters}.  $R^2$ values were calculated for individual runs versus the average of all runs.  Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9931 and standard deviation of 0.0027.  sH1N1 had a mean $R^2$ 0.9848 with a standard deviation of 0.0120.  aH5N1 had a mean $R^2$ of 0.9075 with a standard deviation of 0.1087.

Each run took the calculated viral production and chemokine production rates for the three different strains of influenza as input and reported the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.  Therefore the figures approximate plaque growth over time.

@@

The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).  The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.  By the time T-Cells arrive, the plaque is too large for them to cover effectively.  Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T-Cells to gain ground is too small.  aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{variance}
 \end{center}
\caption{Model results: Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections. IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.  The ten runs were initialized identically for each strain.  Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/mL).} 
 \label{fig:variance}
\end{figure}

@@

T-Cells exit a capillary when they sense inflammatory cytokine and then follow the chemokine gradient in tissue.  We use the chemokine gradient as a proxy for the inflammatory response.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.

To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude. We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e6 and 1e5 $ng/ml$, above which there is no discernible effect on infection kinetics.  Because multiple T-cells in the same area do not increase the rate at which apoptosis is induced, we hypothesize that a critical number of T-cells are required to clear an infection above which there is no additional benefit gained from increasing T-cell presence.  Reference \cite{Gao2003} reports T-cell chemotaxis at concentrations as low as 1 $ng/ml$, therefore we use that sensitivity for each model run.

@@

Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 ineffectual.  We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table~\ref{table:strains}).  Because of this behavior IP-10 was not included in the aH5N1 model runs.

Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).  The lack of both chemokines leads to runaway infections in all three strains.  The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.  IP-10 alone proves to be as effective as the combination of IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.

@@

Early in the infection (up to day 4) the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare. Over time, cells in the plaque's interior die causing active cells to comprise a decreasing proportion of the plaque. T-cells arrive at day 5 and begin to killing the virus-expressing cells. By day 6 a majority of the expressing cells have been eliminated as the plaque becomes dominated by dead cells.  In the H5N1 example the plaque is dense, allowing T-cells to find secreting easily.  Thus the infection can be eliminated.  However, in both H1N1 simulations secreting cells are not eliminated.  In these two cases, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.  T-cells still accumulate, but they arrive at a slower rate than the plaque's rate of growth. This leads to a lower average T-cell killing rate.  Furthermore, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine and for pockets of high chemokine density to decay.  Thus, T-cells, which rely on the chemokine gradient to find infected cells, often lag behind cellular changes in the plaque.  The delayed response and the low proportion of virus secreting cells prevent T-cells from completely clearing the infection.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{cycells}
 \end{center}
\caption{Example simulated sH1N1 infection. Screenshots taken at Day 4, Day 5.5, and Day 7.  The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).  Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).} 
 \label{fig:cycells}
\end{figure}

@@

The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig. \ref{fig:variance}).  This appears to be an artifact of the spatial nature of the ABM.  Why can the immune response create an immediate decrease in the number of pH1N1 infected cells at day 5, only to fail in the same endeavor two days later?   Figure \ref{fig:variance} reports only the number of infected cells, not counting dead cells in the focus of infection (FOI).  Further, T-Cells induce apoptosis in virus-secreting cells, but not in virus-incubating cells.  At day 5.5, the ratio of secreting cells to the total number of cells in the FOI is high (Fig.~\ref{fig:cycells}). By day 7, the ratio of secreting cells to total plaque size is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The greater replication rate of pH1N1 widens the gap (Fig.~\ref{fig:plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to control (but not eliminate) the infection.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  That is, a new viral particle is produced approximately every 200 seconds.  Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit a single infected cell to producing 18 new viral particles.  Without additional immune responses, T-Cells alone cannot contain the pH1N1 infection.  In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing a T-cell to limit a single infected cell to 1.3 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{plaquesize}
 \end{center}
\caption{Comparison of total plaque size (green), number of virus incubating cells (blue), number of virus secreting cells (red), and T-cells (indigo) for the simulated infections of H5N1, sH1N1, and pH1N1.} 
 \label{fig:plaquesize}
\end{figure}

@@

Another advantage of our ABM is the ability to render the model using OpenGL (Fig.~\ref{fig:cycells}).  The ability to see the model in action reveals spatial dynamics that are absent in ODE models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems.  For example, observing the simulation reveals challenges to T-cell detection of infected cells that are not apparent using standard population plots.  By watching the model progress, we observed three key phenomena.  First, because T-cells find infected cells by climbing the chemokine gradient, we see T-cell clustering at local maxima of chemokine concentration.  Thus, we have an explanation for why T-cells do not increase in effectiveness as their numbers increase.  Second, we notice that clumps of T-cells remain in place even after all virus-secreting cells have been killed off.  This is because it takes time for the local chemokine maxima to diffuse and decay to a point where the T-cells can once again climb the gradient to a new maxima.  Finally, we can see that infected cells grow more disperse as the size of the infection grows.  Because T-cells cluster together they become incapable of covering the increasing plaque size.  These spatial observations provide explanations for the pH1N1 resurgence that would be hidden without the visualization tools provided by the ABM.

@@

In both H1N1 strains the T-cell response fails to clear the infection by day 10.  In the case of pandemic H1N1, the infection continues to accelerate after a short period of clearance.  We suggest that this is due to the absence of several innate immune agents that were not included in the model.  We chose to remove most of the innate response in order to better focus on the relative effects of the cytokine signaling and T-cell response.  Although the infections do not clear by day 10, the models show clear differences in the ability of the immune system to deal with influenza strains of varying replication rates.  We see that the viral production rate is a major factor in the strength of the infection.
